Earth's Impact History and the Timing of the Origin of Life

### 2023-04-17 ms.tex
end of introduction

\citet{Benner_2020} argued that the most likely time for the origin of life in an RNA-world scenario would have been $4.36 \pm 0.1$ Ga in the wake of a $2 \times 10^{22}$ kg asteroid impactor which they called “Moneta”. They suggested that iron delivered by Moneta's core would have reacted with impact-vaporized ocean water to generate a reducing Hadean atmosphere conducive to the photochemical generation of essential prebiotic molecules like HCN and HCCCN. This transient period could have been a window of opportunity for the origin of RNA and, ultimately, life.

@@

However, the lunar HSE depletion has explanations that do not require a Moneta impact. Lunar HSEs could have been lost to space during impact-delivery because of the Moon's small gravity \citep{Kraus_2015}. Alternatively, HSEs delivered to the Moon during its $\sim 150$ million-year magma ocean could have been sequestered in the core due to iron sulfide exsolution \citep{Morbidelli_2018,Rubie_2016}. Finally, there is some chance that the Earth's HSEs do not record late impacts because the Moon-forming impact delivered HSEs \citep{Sleep_2016}, or because HSEs are gradual core contributions over time from mantle plumes \citep{Halliday_2023,Mundl_2020}. If indeed Earth's HSEs reflect asteroid bombardment after the Moon formed, then the HSEs can be explained by multiple $\sim 500$ to 2000 km impacts rather than a single Moneta ($\sim 2100$ km) collision.

Recently, \citep{Wogan_2023} considered life-starting impactors significantly smaller than Moneta

@@

\begin{equation}
  \overline{N}_{ij} = N(m_{j-\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}}) - N(m_{j+\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}})
\end{equation}
Next, we sample a Poisson distribution for each $\overline{N}_{ij}$, giving a stochastic number of impacts in each mass and time grid cell which constitutes an impact history. Performing this sampling thousands of times captures many of the possible impact histories. We only keep impact histories that accrete a total mass of $2 \times 10^{22}$ and $6 \times 10^{22}$ kg because this is approximately the mass implied by the highly siderophile elements in Earth's mantle. This range of masses is based on \citet{Day_2015} who used mantle HSEs to suggest that the Earth was impacted by $\sim 3 \times 10^{22}$ to $4.8 \times 10^{22}$ kg of material, but we choose a wider range of masses because HSEs could have been lost to Earth's core or to space during impacts \citep{Marchi_2018}.

@@

>>>>

\citet{Benner_2020} argued that the most likely time for the origin of life in an RNA-world scenario would have been $4.36 \pm 0.1$ Ga in the wake of a $2 \times 10^{22}$ kg ($\sim 2100$ km) asteroid impact which they called Moneta. They suggested that iron delivered by Moneta's core would have reacted with impact-vaporized ocean water to generate a reducing Hadean atmosphere conducive to the photochemical generation of essential prebiotic molecules like HCN. This transient period could have been a window of opportunity for the origin of RNA and, ultimately, life.

@@

However, the lunar HSE depletion has explanations that do not require a Moneta impact. Lunar HSEs could have been lost to space during impact-delivery because of the Moon's small gravity \citep{Kraus_2015}. Alternatively, HSEs delivered to the Moon during its $\sim 150$ million-year magma ocean could have been sequestered in the core due to iron sulfide exsolution \citep{Morbidelli_2018,Rubie_2016}. Finally, there is some chance that the Earth's HSEs do not record late impacts because the Moon-forming impact delivered HSEs \citep{Sleep_2016}, or because HSEs are gradual core contributions over time from mantle plumes \citep{Halliday_2023,Mundl_2020}. If indeed Earth's HSEs reflect asteroid bombardment after the Moon formed, then the HSEs can be explained by multiple $\sim 500$ to 2000 km impacts rather than a single big ($\sim 2100$ km) collision.

Recently, \citep{Wogan_2023} use photochemical models of post-impact atmospheres to show that impacts significantly smaller than Moneta can efficiently produce prebiotic molecules. In their simulations, impactor iron equilibrates with vaporized ocean water to generate atmospheric H$_2$, and CH$_4$ and NH$_3$ form thermochemically as the reducing steam atmosphere cools. Once steam condense to an ocean, subsequent photochemistry of the Titan-like atmosphere produces prebiotic nitriles. Modeling shows that the production of both HCN and HCCCN only occurs when $\mathrm{CH_4} / \mathrm{CO_2} \gtrsim 0.1$ while the atmosphere is hazy. In an optimistic scenario, which includes nickel catalyzed methane production, modeling suggests that $\mathrm{CH_4} / \mathrm{CO_2} > 0.1$ (i.e. significant nitriles) occurs for impacts $> 4 \times 10^{20}$ kg. In a pessimistic modeling case, which assumes only a fraction of impactor iron reacts with the atmosphere \citep{Citron_2022}, a $> 5 \times 10^{21}$ kg impactor is required to produce substantial HCN and HCCCN. Regardless of uncertainty, \citet{Wogan_2023} show that an impactor much smaller than Moneta could spark an origin of life, and that the \citet{Benner_2020} estimated timing for biopoesis could be inaccurate because it is based on Moneta.

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations accounts for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989}. We assume that an ``origin of life impact'' is one that produces significant prebiotic molecules and is not subsequently followed by an ocean-vaporizing impact that destroys the biosphere without rekindling it. By considering the fraction of stochastic impact realizations that do not have an ``origin of life impact'', we also estimate the probability of life beginning on the early Earth if we were to rerun the tape.

@@

\begin{equation}
  \overline{N}_{ij} = N(m_{j-\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}}) - N(m_{j+\frac{1}{2}},t_{i-\frac{1}{2}},t_{i+\frac{1}{2}})
\end{equation}
Next, we sample a Poisson distribution for each $\overline{N}_{ij}$, giving a stochastic number of impacts in each mass and time grid cell which constitutes an impact history. Performing this sampling thousands of times captures many of the possible impact histories. We only keep impact histories that accrete a total mass of $2 \times 10^{22}$ and $6 \times 10^{22}$ kg because this is approximately the mass implied by the HSEs in Earth's mantle. This range of masses is based on \citet{Day_2015} who used mantle HSEs to suggest that the Earth was impacted by $\sim 3 \times 10^{22}$ to $4.8 \times 10^{22}$ kg of material, but we choose a wider range of masses because HSEs could have been lost to Earth's core or to space during impacts \citep{Marchi_2018}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{figures/example_impact_histories.pdf}
  \caption{Caption}
  \label{fig:example_impact_histories}
\end{figure}
